D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-16 ootavacuous.tex
December 15 self-review plus narrate JVM litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
The Indefatigible TBD
}
\date{November 27, 2023 (Pre-Tokyo)}
\maketitle{}

@@

If additional inter-thread information is available, the corresponding
constraints may be applied.
For example, if an implementation determines that the values of a pair
of objects are always equal at any time that a given thread might access
them, one could imagine a sufficiently brave implementation substituting
a load from one for a load from the other.

However, if all of the inter-thread relaxed accesses are volatile,
as they are in the C language, then all of the inter-thread traffic is
observed behavior.
In this case, even omniscient oracular implementations are required
to confine their analyses to within a single thread.

@@

@@

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or non-semantic on the other.
While taking this approach sacrifices significant generality, it has the
significant benefit of greatly reducing the cost of dependency analysis,
potentially all the way down to zero.

@@

Rather than look at how to prevent an implementation from forming
an OOTA cycle, this section instead looks at how to form one.
This will lead to the conclusion that a correct C++ implementation
cannot form an OOTA cycle.

@@

But it is also the case that an OOTA cycle must have at least one link
that goes backwards in time.
Given that the inter-thread links are temporal, the only possible
candidate atemporal links so are those confined to a given thread.
This in turn requires that a store that depends on an earlier load
be executed earlier in time than that load.
The two ways that this can happen are when the implementation:

\begin{enumerate}
\item	Is able to determine (prove) the location and value of the store.
\item	Guesses the location and value of the store and has some means
	to take corrective action should any guess prove incorrect.
\end{enumerate}

@@

Implementation can sometimes prove the location and value of a given
store.
For example, if the value loaded is multiplied by zero, the result
is known to be zero, so that the store can proceed prior to the load.
But in this case, there is no perturbation of the value returned from
the load that can affect the value stored.
Therefore, by P0442R0, that store cannot possibly take part in an
OOTA cycle.

@@

Hardware speculative execution is commonplace, and permits the
hardware to guess at values, squashing the speculation if any
of the guesses prove incorrect.
But in this case, a given store is not committed until all guesses that
this store depends on have been confirmed, and thus no uncommitted store
is visible to any other thread.
Therefore, hardware speculation cannot result in OOTA cycles, and
by design, even on weakly ordered
systems~\cite{ARMv7A:2010,ARMv8A:2017,PowerISA2.07-2013}.
This same constraint applies to software, which might use compare-and-swap
loops or hardware transactional memory to confirm guesses.

@@

This paper has added temporal reasoning to OOTA analysis of the C++
memory model.
This work shows that correct C++ implementations are inherently incapable
of producing OOTA cycles.

Therefore, if your C++ implementation produces OOTA cycles, you have
bugs that would be considered bugs even if OOTA cycles were allowed.
And you need to fix those bugs, regardless of whether they are in your
C++ implementation, your hardware, or your alternate universe.

@@

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
However, the \co{herd7} tool does not find this litmus test's cycle, presumably
because this tool does not do the deep inter-thread program analysis
required to find it.

@@

@@

@@

@@

Listing~\ref{lst:Causality Test Case 4}
shows causality test case 4, for which the \co{r1 == r2 == 1} result
is to be forbidden.
The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.

@@

@@

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.

@@

@@

Listing~\ref{lst:Causality Test Case 6}
shows causality test case 6, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.

@@

@@

@@

@@

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
initialized to the constant \co{1}, it does find the cycle and refrains
from flagging it as an OOTA cycle.

@@

@@

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
initialized to the constant \co{1}, it does find the cycle and refrains
from flagging it as an OOTA cycle.

@@

Listing~\ref{lst:Causality Test Case 10}
shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The \co{herd7} tool nevertheless reports this result and does not flag it as
an OOTA cycle.

@@

@@

@@

JMM Causality Test Case 12 uses arrays, which are not yet supported
by the \co{herd7} tool.

@@

Listing~\ref{lst:Causality Test Case 13}
shows causality test case 13, for which the \co{r1 == r2 == 1}
result is to be forbidden.
The \co{herd7} tool nevertheless reports this result and does not flag it as
an OOTA cycle.

@@

@@

Listing~\ref{lst:Causality Test Case 14}
shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
result is to be forbidden.
The \co{herd7} tool does not report this result.

@@

@@

Listing~\ref{lst:Causality Test Case 15}
shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
result is to be forbidden.
The \co{herd7} tool does not report this result.

@@

@@

Listing~\ref{lst:Causality Test Case 16}
shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given the differences between Java on the one hand and C and C++
on the other.

@@

Listing~\ref{lst:Causality Test Case 17}
shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given the differences between Java on the one hand and C and C++
on the other.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S17}.

@@

Listing~\ref{lst:Causality Test Case 18}
shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given the differences between Java on the one hand and C and C++
on the other.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S17}.

@@

Listing~\ref{lst:Causality Test Case 19}
shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given the differences between Java on the one hand and C and C++
on the other.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S8}.

@@

@@

Listing~\ref{lst:Causality Test Case 20}
shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given the differences between Java on the one hand and C and C++
on the other.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S8}.

@@

\begin{listing}[tbp]
\begin{verbatim}
 1 std::atomic<int> y;
 2 template <typename T>
 3 void t1() {
 4   long long r1 = x.load(std::memory_order_relaxed);
 5   if (r1 % 4 == 0)
 6     y.store(1, std::memory_order_relaxed);
 7   __asm__ __volatile__("": : :"mightnotreturn");
 8   *(T*)r1 = *(T*)r1;
 9 }
\end{verbatim}
\caption{Alignment-Based Undefined Behavior Fix}
\label{lst:Alignment-Based Undefined Behavior Fix}
\end{listing}

However, in that case, the developer is stepping outside the bounds
of the standard.
It would be good to have some way for the developer to get the job
done without confusing the compiler, thus separating OOTA-cycle and
UB concerns.
One approach is to have a \co{mightnotreturn} clobber similar to the
\co{memory} clobber for GCC asms.
Then this example could be rewritten as shown in
Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
with the asm on line~7 preventing the compiler from assuming that line~8
is guaranteed to be executed, thus preventing the compiler from
applying any lessons that it might learn from line~8 to line~5.

@@

This \co{mightnotreturn} clobber is merely an example.
There might well be better ways to prevent UB from propagating backwards.
Other options being considered include a call to an empty function that
is invisible to the compiler and a GCC goto asm that the compiler cannot
prove falls through.

A \co{std::observable()} function that would block backwards-propagating
UB has been proposed to the C++ standards
committee~\cite{DavisHerring2021P1494R2}.

@@

This section attempts to apply the lessons
of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
Goldblatt's unreachability-UB example.  Inserting the asm at line~12
as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
avoids confusing the compiler into assuming that the value of \co{local}
must necessarily be zero.

\begin{listing}[tbp]
\begin{verbatim}
 1 // UB if this function is ever called.
 2 [[noreturn]]
 3 inline void unreachable() {
 4   return;
 5 }
 6 void t1() {
 7   r1 = x.load(std::memory_order_relaxed);
 8   int local = (r1 & 1);
 9   if (local == 0)
10     y.store(1, std::memory_order_relaxed);
11   if (local != 0) {
12     __asm__ __volatile__("": : :"mightnotreturn");
13     // If we get here, UB.
14     unreachable();
15   }
16 }
\end{verbatim}
\caption{Unreachability-Based Undefined Behavior Fix}
\label{lst:Unreachability-Based Undefined Behavior Fix}
\end{listing}

By the same reasoning, Goldblatt's \co{misbhave} example that
also involves unreachability can also be addressed through insertion
of the asm.

@@

% Additional snark:
% "This work reduces the problem of OOTA analysis to that of sequential
%	program analysis.  If you expect better verification of concurrent
%	code than of sequential code, you are in an intellectual state
%	of sin."
% "Forcing relaxed loads to be ordered before subsequent relaxed stores
%	is an expensive no-op."
% "So you don't want to learn concurrency?  Then I can only suggest
%	that you either move further up the stack or take early
%	retirement."

>>>>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
The Indefatigible TBD
}
\date{December 15, 2023 (Pre-Tokyo)}
\maketitle{}

@@

If additional inter-thread information is available, the corresponding
constraints may be applied.
For example, if an implementation proves that the values of a pair of
objects are always equal at any time that a given thread might access
them, that implementation could behave as if the that thread's code had
substituted a load from one for a load from the other.

Note that if all of the inter-thread relaxed accesses are volatile,
as they are in the C language, then all of the inter-thread traffic is
observed behavior.
In this case, even omniscient oracular implementations are required
to confine their analyses to within a single thread.

@@

Another question is how to prove that a given optimization is valid.
This has been a hard question in the past, and is likely to remain so
both for sequential and concurrent programs.
The consequences of using an invalid optimization can be quite severe,
and the evidence thus far indicates that OOTA cycles are not a bit issue.
Nevertheless, a proof that the same value is always stored implies a
proof that there is no semantic dependency to that relaxed atomic store,
and thus no chance of that store participating in an OOTA cycle.

@@

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or non-semantic on the other,
whether by working with that implementation's intermediate representations,
binaries produced by that implementation, or by tracing the program's
execution.
While taking this approach sacrifices significant generality, it has the
significant benefit of greatly reducing the cost of dependency analysis,
potentially all the way down to zero.

@@

Rather than look at how to prevent an implementation from forming
an OOTA cycle, this section instead looks at how to form one.
This will lead to the conclusion that a correct real-world C++
implementation cannot form an OOTA cycle.

@@

But it is also the case that an OOTA cycle must have at least one link
that goes backwards in time.
Given that the inter-thread links are temporal with a give store preceding
any load returning the value stored, the only possible candidate atemporal
links are those confined to a given thread.
This in turn requires that a store that depends on an earlier load
be executed earlier in time than that load.
The two ways that this can happen are when the implementation:

\begin{enumerate}
\item	Is able to prove the location and value of the store.
\item	Guesses the location and value of the store and has some means
	to take corrective action should any guess prove incorrect.
\end{enumerate}

@@

Implementation can sometimes prove the location and value of a given
store.
For example, if the value loaded is multiplied by zero, the result
is known to be zero, so that the store can proceed prior to the load.
But in this case, there is no way that the value returned from
the load that can affect the value stored.
Therefore, that store cannot possibly take part in an OOTA cycle.

@@

Hardware speculative execution is commonplace, and permits the
hardware to guess at values, squashing the speculation if any
of the guesses prove incorrect.
But in this case, a given store is not committed until all guesses that
this store depends on have been confirmed, and thus no uncommitted store
is visible to any other thread.
Therefore, any hardware speculation that results in OOTA cycles is doomed
to be squashed an discarded, and by design, even on weakly ordered
systems~\cite{ARMv7A:2010,ARMv8A:2017,PowerISA2.07-2013}.
This same constraint applies to software, which might use compare-and-swap
loops or hardware transactional memory to confirm guesses.

@@

This paper has added temporal reasoning to OOTA analysis of the C++
memory model.
This work shows that correct real-world C++ implementations are inherently
incapable of producing OOTA cycles.

Therefore, if your C++ implementation produces OOTA cycles, you have bugs.
And you need to fix those bugs, regardless of whether they are in your
C++ implementation, your hardware, or your alternate universe.

@@

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
The \co{herd7} tool finds this litmus test's cycle, even without the
inter-thread analysis, because C++ does not respect control dependencies
to relaxed atomic stores.

@@

Implementations that can prove that the value of \co{x} is always
non-negative can also prove that there is no semantic dependency between
\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
that is, no matter which of the possible values is loaded from \co{x},
the value 1 will always be stored to \co{y}.

Less-omniscient implementations, including those whose code analysis
is restricted to a thread, would need to act as if there was in fact a
semantic dependency.

Even omniscient implementations will observe a semantic dependency from
\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.

Either way, there is no OOTA cycle, with omniscient implementations
having no cycle at all, and other implementations retaining the
non-semantic dependency.

@@

Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
an implementation can act as if the source code loaded from \co{x} only
once and assigned the returned value to both \co{r1} and \co{r2}.\footnote{
	But not in C, where relaxed atomic operations are volatile,
	and thus observable behavior.}
The implementation could then prove that the condition on line~10
would always evaluate to \co{true}, and act as if the source code had
unconditionally executed the relaxed store to \co{y} on line~11.
The implementation could then act as if the resulting relaxed accesses
to \co{x} and \co{y} appeared in reverse order, at which point
\co{P0()}'s store to \co{y} could be read by \co{P1()} load so that
\co{P1()} stored 1 to \co{x}, thus resulting in \co{P1()} loading
that 1 from \co{x}.

Note that \co{P1()} has a trivial semantic dependency from its load
from \co{y} on line~15 to its store to \co{x} on line~16.

If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
otherwise the preserved non-semantic dependency prevents an
OOTA cycle from forming on correct real-world implementations.

@@

The analysis from Section~\ref{app:Causality Test Case 2}
applies here, including the difference in volatility between C and C++
atomics.
Although the addition of \co{P2} allows sequentially consistent executions
where \co{P0()}'s two loads from \co{x} might return different values,
it does not change the fact that those two loads can be merged.

@@

Listing~\ref{lst:Causality Test Case 4}
shows causality test case 4, for which the \co{r1 == r2 == 1} result
is to be forbidden.
The \co{herd7} tool finds this OOTA cycle because C++ does not respect
data dependencies, and flags it as such using \co{S8}.

@@

Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
from their relaxed atomic loads to their relaxed atomic stores, which
prevents this OOTA cycle on real-world C++ implementations.

@@

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The \co{herd7} tool finds this OOTA cycle because C++ does not respect
data dependencies, and flags it as such using \co{S8}.

@@

Each of \co{P0()}, \co{P1()}, and \co{P3()} have trivial semantic
dependencies from their relaxed atomic loads to their relaxed
atomic stores, which prevents any OOTA cycle from forming in correct
real-world C++ implementations.
\co{P2()} has no loads, so cannot participate in an OOTA cycle in
any case.

@@

Listing~\ref{lst:Causality Test Case 6}
shows causality test case 6, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool finds this cycle, and does not flag it as an OOTA
cycle because C++ does not respect control dependencies to relaxed
atomic stores.

@@

A C++ implementation that can prove that \co{y} is always either
zero or one can also prove that there is no semantic dependency
from \co{P1()}'s load on line~14 to its stores on lines~16 and~29.
Such an implementation could act as if lines~15-20 had been replaced
by a single store to \co{x}, allowing that store to execute first,
followed by lines~8, 10, and~14, in that order, satisfying the
\co{exists} clause on line~23.

Less-omniscient implementations will preserve the non-semantic
dependency, preventing any OOTA loop from forming in a correct real-world C++
implementation.

@@

Because all accesses are relaxed atomics, this test case has everything
to do with reordering and nothing to do with OOTA cycles.
The outcome happens when lines~17, 10, 11, 15, 16 and~9 execute in that
order.
Note that this ordering preserves the semantic dependencies from
lines~10-11 and~15-16, and can thus happen on correct real-world
C++ implementations.

@@

Omniscient C++ implementations might note that the computation on line~9
can only have the value one when the load from \co{x} returns either
zero or one, and both \co{x} and \co{y} are initialized to zero.
Such an implementation act as if line~10's \co{r2} had instead been
the value~1,
Then there would be no semantic dependency between \co{P0()}'s load
and store, and thus no OOTA cycle in a correct real-world C++
implementation.

A less-capable implementation would leave the non-semantic dependency
in place, also preventing any OOTA cycle in a correct real-world C++
implementation.

@@

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} after the other two processes, and C++ is not
constrained to do this.

@@

Both processes have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} before the other two processes, and C++ is not
constrained to do this.

@@

Listing~\ref{lst:Causality Test Case 10}
shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
Because C++ does not respect control dependencies, the \co{herd7} tool
nevertheless reports this result and does not flag it as an OOTA cycle.

@@

However, each of \co{P0()}, \co{P1()}, and \co{P3()} have straightforward
semantic dependencies between their respective loads and stores, and
thus cannot form an OOTA cycle when running on a correct real-world C++
implementation.
For its part, \co{P2()} has no semantic dependency, so it cannot be part
of an OOTA cycle when running on a correct real-world C++ implementation.

@@

This test case has nothing to do with OOTA cycles.
The result may be obtained by executing the statements in the following
order: line~20, line~12, line~13, line~18, line~19, line~10, line~11,
and finally line~17.

@@

JMM Causality Test Case 12 in
Figure~\ref{lst:Causality Test Case 12}
uses arrays, which are not yet supported by the \co{herd7} tool.

\begin{listing}[tbp]
@@ DisplayLitmus litmus/oota-causality-12.litmus @@
\caption{Causality Test Case 12}
\label{lst:Causality Test Case 12}
\end{listing}

An omniscient C++ implementation could note that only \co{P0()} accesses
array \co{a[]} and act as if the code was simpler, but there would
still be semantic dependencies between \co{P0()}'s and \co{P1()}s
relaxed atomic loads and stores involving \co{x} and \co{y}.
Those dependencies prevent an OOTA cycle from forming on a correct
real-world C++ implementation, omniscient or not.

@@

Listing~\ref{lst:Causality Test Case 13}
shows causality test case 13, for which the \co{r1 == r2 == 1}
result is to be forbidden.
But because C++ does not respect control dependencies, \co{herd7} tool
reports this result and does not flag it as an OOTA cycle.

@@

There are straightforward semantic dependencies between \co{P0()}'s and
\co{P1()}s relaxed atomic loads and stores involving \co{x} and \co{y}.
Those dependencies prevent an OOTA cycle from forming on a correct
real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 14}
shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
result is to be forbidden.
The \co{herd7} tool nevertheless reports this result because C++ does not
respect control dependencies for atomic stores.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
line~13, line~18 (so \co{r3 == 1)}.
The sum of \co{r2} and \co{r3} is non-zero, justifying line~21's
store to \co{x}.

However, correct real-world C++ implementations would have semantic
dependencies between \co{P0()}'s line~9 and lines~11 and~13, and also
between \co{P1()}'s lines~17 and~18 and line~21, preventing any
OOTA cycle from forming.

@@

Listing~\ref{lst:Causality Test Case 15}
shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
result is to be forbidden.
The \co{herd7} tool nevertheless reports this result because C++ does not
respect control dependencies for atomic stores.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
and finally line~24 (so \co{r3 == 1}).
The sum of \co{r2} and \co{r3} is non-zero, justifying line~27's
store to \co{u}.

However, correct real-world C++ implementations would have semantic
dependencies between \co{P0()}'s lines~10 and~13 and lines~17 and~19, and also
between \co{P1()}'s lines~23 and~24 and line~27, preventing any
OOTA cycle from forming.
Having no load, \co{P2()} cannot participate in an OOTA cycle.

@@

Listing~\ref{lst:Causality Test Case 16}
shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.

@@

Listing~\ref{lst:Causality Test Case 17}
shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S17}.

@@

Listing~\ref{lst:Causality Test Case 18}
shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S17}.

@@

Listing~\ref{lst:Causality Test Case 19}
shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S8}.

It is interesting that this \co{S8} value does not propagate to
\co{r3}.

@@

All processes have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 20}
shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S8}.

It is interesting that this \co{S8} value does not propagate to
\co{r3}.

@@

\begin{listing}[tbp]
\begin{verbatim}
 1 std::atomic<int> y;
 2 template <typename T>
 3 void t1() {
 4   long long r1 = x.load(std::memory_order_relaxed);
 5   if (r1 % 4 == 0)
 6     y.store(1, std::memory_order_relaxed);
 7   std::observable();
 8   *(T*)r1 = *(T*)r1;
 9 }
\end{verbatim}
\caption{Alignment-Based Undefined Behavior Fix}
\label{lst:Alignment-Based Undefined Behavior Fix}
\end{listing}

However, in that case, the developer is stepping outside the bounds
of the standard.
It would be good to have some way for the developer to get the job
done without confusing the compiler, thus separating OOTA-cycle and
UB concerns.
One approach is to use the \co{std::observable()} function
that has been proposed to the C++ standards
committee~\cite{DavisHerring2021P1494R2}.
This function would blocks backwards-propagating UB.
Then this example could be rewritten as shown in
Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
with the asm on line~7 preventing the compiler from assuming that line~8
is guaranteed to be executed, thus preventing the compiler from
applying any lessons that it might learn from line~8 to line~5.

@@

There are a number of ways to implement \co{std::observable()},
for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
function that is invisible to the compiler, and a GCC goto asm that the
compiler cannot prove falls through.

@@

This section attempts to apply the lessons
of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
Goldblatt's unreachability-UB example.

Inserting \co{std::observable()} at line~12
as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
avoids confusing the compiler into assuming that the value of \co{local}
must necessarily be zero.

\begin{listing}[tbp]
\begin{verbatim}
 1 // UB if this function is ever called.
 2 [[noreturn]]
 3 inline void unreachable() {
 4   return;
 5 }
 6 void t1() {
 7   r1 = x.load(std::memory_order_relaxed);
 8   int local = (r1 & 1);
 9   if (local == 0)
10     y.store(1, std::memory_order_relaxed);
11   if (local != 0) {
12     std::observable();
13     // If we get here, UB.
14     unreachable();
15   }
16 }
\end{verbatim}
\caption{Unreachability-Based Undefined Behavior Fix}
\label{lst:Unreachability-Based Undefined Behavior Fix}
\end{listing}

By the same reasoning, Goldblatt's \co{misbhave} example that
also involves unreachability can also be addressed through insertion
of the proposed \co{std::observable()}.

@@

% Additional snark:
% "This work reduces the problem of OOTA analysis to that of sequential
%	program analysis.  If you expect better verification of concurrent
%	code than of sequential code, you are in an intellectual state
%	of sin."
% "Forcing relaxed loads to be ordered before subsequent relaxed stores
%	is an expensive no-op."
% "So you don't want to learn concurrency?  Then I can only suggest
%	that you either move further up the stack or take early
%	retirement."  Note that that C++ committee's concurrency
%	expertise has grown beyond recognition over the past 15 years.
